\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-12-14 doc/paper-jss/sdmTMB-paper.Rnw
Copy editing; add JT

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex")

@@

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
Sean C. Anderson~\orcidlink{0000-0001-9563-1937}\\Pacific Biological
Station \And Eric J. Ward~\orcidlink{0000-0002-4359-0296}\\Northwest
Fisheries Science Center \AND Philina A.
English~\orcidlink{0000-0003-2992-6782}\\Pacific Biological
Station \And Lewis A. K. Barnett~\orcidlink{0000-0002-9381-8375}\\Alaska
Fisheries Science Center
}
\title{\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and
User-Friendly Generalized Linear Mixed Effects Models with Spatial and
Spatiotemporal Random Fields}

\Plainauthor{Sean C. Anderson, Eric J. Ward, Philina A. English, Lewis
A. K. Barnett}
\Plaintitle{sdmTMB: An R Package for Fast, Flexible, and Accessible
Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal
Random Fields}
\Shorttitle{\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with
\pkg{TMB}}

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent Gaussian Markov random fields using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

\Keywords{Gaussian Markov random fields (GMRF), generalized linear mixed effects
models (GLMM), INLA, SPDE, species distribution modelling, R package,
spatio-temporal, spatial-temporal, Template Model Builder}

\Plainkeywords{Gaussian Markov random fields (GMRF), generalized linear mixed
effects models (GLMM), INLA, SPDE, species distribution modelling, R package,
spatio-temporal, spatial-temporal, Template Model Builder}

@@

  }

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep[e.g.,][]{thorson2019, osgood-zimmerman2021}.

Software packages designed for specifying statistical models that incorporate the SPDE, such as \INLA\ and \TMB, are flexible and powerful but can be challenging for many applied researchers.
For example, \TMB\ requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

A basic GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
\]
constrained by some function that defines the rate at which spatial covariance decays with distance.

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices, thorson2018comp}.

@@

In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develops the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} contains an introduction to the SPDE and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provides a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally fit as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
% A19
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristkristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

Additional utilities that require heavier package dependencies (such as \INLA\ and \pkg{rstan}) are maintained in the \pkg{sdmTMBextra} package at\\
\url{https://github.com/pbs-assess/sdmTMBextra}

>>>>

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)

@@

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
Sean C. Anderson~\orcidlink{0000-0001-9563-1937}\\Pacific Biological
Station \And
Eric J. Ward~\orcidlink{0000-0002-4359-0296}\\Northwest
Fisheries\\Science Center \And
Philina A. English~\orcidlink{0000-0003-2992-6782}\\Pacific Biological
Station \AND
Lewis A. K. Barnett~\orcidlink{0000-0002-9381-8375}\\Alaska
Fisheries Science Center \And
James T. Thorson~\orcidlink{0000-0001-7415-1010}\\Alaska
Fisheries Science Center\\
}
\title{\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and
User-Friendly Generalized Linear Mixed Effects Models with Spatial and
Spatiotemporal Random Fields}

\Plainauthor{Sean C. Anderson, Eric J. Ward, Philina A. English, Lewis
A. K. Barnett, James T. Thorson}
\Plaintitle{sdmTMB: An R Package for Fast, Flexible, and Accessible
Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal
Random Fields}
\Shorttitle{\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with
\pkg{TMB}}

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

\Keywords{Gaussian Markov random field (GMRF), GLMM, SPDE, species distribution modelling, R package, spatial-temporal, Template Model Builder (TMB)}

\Plainkeywords{Gaussian Markov random field (GMRF), GLMM, SPDE, species distribution modelling, R package, spatial-temporal, Template Model Builder (TMB)}

@@

    James T. Thorson\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and
    Atmospheric Administration, Seattle, WA, USA
  }

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

Such models quickly become computationally challenging due to the need to invert large matrices to account for covariation among data.
Many solutions have been proposed, such as predictive processes \citep{banerjee2008, latimer2009}, the stochastic partial differential equation (SPDE) approximation to GRFs \citep{lindgren2011}, and nearest-neighbour Gaussian processes \citep{datta2016, finley2022}.
These approaches aim to minimize the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian \citep{gelfand2017} or maximum likelihood methods.
This can greatly improve computational efficiency \citep[e.g.,][]{heaton2019}.
The SPDE approach has been widely adopted, especially via the \INLA\ package \citep{rue2009, lindgren2011, lindgren2015} and an implementation in \TMB\ \citep[Template Model Builder,][]{kristensen2016} that relies on \INLA\ to create input matrices \citep{thorson2015ecology, thorson2015ices, thorson2019, osgood-zimmerman2023}.

Software packages designed for specifying statistical models that incorporate the SPDE, such as \INLA\ and \TMB, are flexible and powerful but can be challenging for many applied researchers.
For example, \TMB\ requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
While packages such as \pkg{lme4} \citep{bates2015} and \pkg{glmmTMB} \citep{brooks2017} let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---they lack built-in SPDE or spatiotemporal functionality.
Packages such as \pkg{VAST} \citep{thorson2019} and \pkg{inlabru} \citep{bachl2019} are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, lack some functionality, or may be challenging to learn for some users.

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

A GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_{\bm{s}}$.
That mean $\mu_{\bm{s}}$ is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta_{\bm{s}}$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
\]
constrained by some function that defines the rate at which spatial covariance decays with distance.

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{haskard2007, lindgren2011, thorson2015ices}.

@@

In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develop the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
% A19
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
In this paragraph, we include relevant \fct{predict} arguments in parentheses.
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
which can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at\\
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

Additional utilities, which require heavier package dependencies (such as \INLA\ and \pkg{rstan}) and are used by only a subset of users, are maintained in the \pkg{sdmTMBextra} package at\\
\url{https://github.com/pbs-assess/sdmTMBextra}
